\Ace: Growing a Statically-Typed Language Inside a Python

### 2013-11-21 ace-pldi14/ace-pldi14.tex
Added mbeddr reference to PLDI14

\section{Expressiveness}\label{examples}
Thus far, we have focused mainly on the OpenCL target and shown examples of fairly low-level active types: those that implement OpenCL's primitives (e.g. \verb|clx.Ptr|), extend them in simple but convenient ways (e.g. \verb|clx.Cplx|) and those that make interactive execution across language boundaries safe and convenient (e.g. \verb|clx.NPArray|). 
Ace was first conceived to answer the question: \emph{can we build a statically-typed language with the semantics of a C but the syntax and ease-of-use of a Python?} We submit that these examples (along with a large-scale simulator framework built using them, which we do not have space to detail here) have answered this question in the affirmative.

@@

Rule Injection && Qi, Typed Racket \cite{TypedScheme2008}, Xroma \cite{activelibraries} && some && \Circle && \CIRCLE && \Circle && \Circle \\

@@

Libraries containing compile-time logic have previously been called {\it active libraries} \cite{activelibraries}. A number of  projects, such as Blitz++, have taken advantage of C++ template metaprogramming to implement domain-specific optimizations \cite{veldhuizen2000blitz++}. Others have chosen custom metalanguages (e.g. Xroma \cite{activelibraries}). In Ace, we replace these brittle mini-languages with a general-purpose language, Python, and significantly expand the notion of active libraries by consideration of types, base semantics and target languages as values (in this case, objects) in the compile-time metalanguage.

@@

There are several other language-internal extension mechanisms that have been described in the literature. These differ from our mechanism in one of the following ways, summarized in Fig. \ref{relatedtable}:
\begin{itemize}
\item Our mechanism is itself implemented as a library, rather than as a language extension.
\item Our mechanism reuses an existing language's syntax, rather than offering syntax extension support. As we saw in, for example, Listing \ref{datatypest}, this can still be reasonably natural, and allows the language to benefit from a variety of existing tools.
\item Our mechanism permits true type system extensions. New types are not merely aliases, nor must their rules be admissible in some base type system.% Term rewriting and macro systems do not handle types at all.
\item As described in Sec. \ref{safety}, our mechanism permits compositional reasoning. Techniques that allow global extensions to the syntax or semantics (e.g. SugarJ or rule injection systems like Xroma or Typed Racket) allow extensions so much control that they can interfere with one another or make reasoning  difficult.
\item Our mechanism allows users to control the target language of compilation from within libraries. Many other mechanisms are based fundamentally on term-to-term rewriting.
\end{itemize}

>>>>

\section{Expressiveness}\label{examples}
Thus far, we have focused mainly on the OpenCL target and shown examples of fairly low-level active types: those that implement OpenCL's primitives (e.g. \verb|clx.Ptr|), extend them in simple but convenient ways (e.g. \verb|clx.Cplx|) and those that make interactive execution across language boundaries safe and convenient (e.g. \verb|clx.NPArray|). 
Ace was first conceived to answer the question: \emph{can we build a statically-typed language with the semantics of a C but the syntax and ease-of-use of a Python?} We submit that the work described above has answered this question in the affirmative.

@@

Rule Injection && Racket \cite{TypedScheme2008}, Xroma \cite{activelibraries}, mbeddr \cite{mbeddr} &&  \cite{TypedScheme2008} && \cite{mbeddr} && \CIRCLE && \Circle && \Circle \\

@@

Libraries containing compile-time logic have previously been called {\it active libraries} \cite{activelibraries}. A number of  projects, such as Blitz++, have taken advantage of C++ template metaprogramming to implement domain-specific optimizations \cite{veldhuizen2000blitz++}. Others have chosen custom metalanguages (e.g. Xroma \cite{activelibraries}, mbeddr \cite{mbeddr}). In Ace, we replace these brittle mini-languages with a general-purpose language, Python, and significantly expand the notion of active libraries by consideration of types, base semantics and target languages as values (in this case, objects) in the compile-time metalanguage.

@@

There are several other language-internal extension mechanisms that have been described in the literature. These differ from our mechanism in one of the following ways, summarized in Fig. \ref{relatedtable}:
\begin{itemize}
\item Our mechanism is itself implemented as a library, rather than as a language extension. Only typed LISPs are similar.
\item Our mechanism reuses an existing language's syntax, rather than offering syntax extension support. As we saw in, for example, Listing \ref{datatypest}, this can still be reasonably natural, and allows the language to benefit from a variety of existing tools.
\item Our mechanism permits true type system extensions. New types are not merely aliases, nor must their rules be admissible in some base type system.% Term rewriting and macro systems do not handle types at all.
\item As described in Sec. \ref{safety}, our mechanism permits compositional reasoning. Techniques that allow global extensions to the syntax or semantics (e.g. SugarJ or rule injection systems like Xroma, Typed Racket (and other typed LISPs) and mbeddr) allow extensions so much control that they can interfere.
\item Our mechanism allows users to control the target language of compilation from within libraries. Many other mechanisms are based fundamentally on term-to-term rewriting.
\end{itemize}
